Thesis Title

### 2022-02-04 chapters/introduction.tex
Continue updating introduction

In addition to the emergence of virulent pathogens, the rise of antimicrobial resistance (AMR) poses a major threat to human health around the world. In 2019 there were an estimated 4·95 million deaths associated with bacterial AMR \citep{murray_global_2022}. In 2017, the WHO released The Global Priority Pathogens (GPP) list \citep{organization_prioritization_2017} to guide discovery, research and development of new antibiotics for drug-resistant bacterial infections. Besides tuberculosis, the global priority due to being the most common and lethal airborne AMR disease worldwide today, responsible for 250 000 deaths each year, it includes 12 groups of pathogens in three priority categories:

@@

\subsection{Current standards for diagnostic and public health in clinical microbiology}

The past few decades have seen a major revolution in the operation of microbial laboratories, driven by the development of molecular technologies and ways to make these accessible, namely amplification-based polymerase chain reaction (PCR), matrix-assisted laser desorption/ionisation - time of light (MALDI-TOF), DNA-microarray-based hybridisation technology and T2 Magnetic Resonance (T2MR). These are used in conjunction with traditional techniques such as microscopy, culture and serology, not fully replacing them. Application of these methods differs by suspected infection type: bacterial, viral, fungal or parasitic.

@@

For patients with bacterial infections, the crucial steps are (1) to grow an isolate from a specimen, (2) identify its species, and (3) determine its pathogenic potential and test its susceptibility to antimicrobial drugs  \citep{didelot_transforming_2012}. Together this information facilitates the specific and rational treatment of patients. For public health purposes, knowledge also needs to be gained about (4) the relatedness of the pathogen to other strains of the same species to investigate transmission routes and enable the recognition of outbreaks \citep{foxman_choosing_2005}. Traditionally, all of these steps rely on the isolation of the pathogen through culture, followed by a multi-step process that can take days to weeks before obtaining results, depending on the fastidiousness of the organism and if it can be cultured.

@@

In the early 2000s, Multilocus sequence typing (MLST) was proposed as a portable, universal, and definitive method for characterising bacteria \citep{maiden_multilocus_2006}. Instead of enzyme restriction of bacteria DNA, separation of the restricted DNA bands using a pulsed-field electrophoresis chamber, followed by clonal assignment of bacteria based on PFGE banding patterns, MLST relies on the amplification through PCR sequences of internal fragments of house-keeping genes (usually 5 to 7), approximately 450-500 basepairs (bp) in size, followed by its the sequence, usually my Sanger methods. For each house-keeping gene, the different sequences present within a bacterial species are assigned as distinct alleles and, for each isolate, the alleles at each of the (usually) seven loci define the allelic profile or sequence type \citep{larsen_multilocus_2012}. As with PFGE, different schemes, defining what house-keeping gene fragments are used, are available depending of the species. Unlike PFGE, the provision of freely accessible, curated databases of MLST nucleotide sequence data enables the direct comparison of bacterial isolates, providing the basis of a common language for bacterial typing \citep{maiden_multilocus_2006}. MLST schemes for over 100 organisms having been published and made freely available (\url{https://pubmlst.org/organisms}, \cite{jolley_open-access_2018})

@@

Direct detection methods negate the need for virus propagation, detecting the virus directly from the suspect source through nucleic acid and immunological methods. PCR and reverse transcription-PCR are widely applied methods for the detection of both DNA and RNA viruses, respectively, driven by an increased awareness of the clinical value of, and demand for, prompt information about viral loads, viral sequence data, and potential antiviral resistance information. %CITATION
Real-time quantitative PCR (qPCR) remains the front line tool in aetiological diagnosis, measuring the production of the target amplicon throughout the reaction and providing quantitative results with high specificity and sensibility, albeit with a significant cost due to sophisticated apparatus despite high-throughput systems being widely established \citep{cassedy_virus_2021}. Immunoassays employ singular-epitope specificity antibodies as the primary means to detect viruses within a sample, and provides as a much more cost-efficient alternative to nucleic acid detection \citep{cassedy_virus_2021}. One major application is seroprevalence assays, an essential technique for identifying patients who have been exposed to a virus (historical exposure), detect asymptomatic infection or evaluate vaccine efficacy  \citep{chan_determining_2021, bobrovitz_global_2021}. Lateral flow immunoassays (LFIA) are extensively used for detecting virus-associated protein directly from source through labelled antibodies binding to their cognate antigens, usually read by way of a colour change at a test line. Besides being very cost effective, LFAIs have a turnaround time of minutes and the colour change can be observed with the naked eye, therefore facilitating rapid diagnosis but its results are limited to semi-quantitative and it does not typically achieve sensitivity comparable to nucleic-acid detection \citep{estrela_lateral_2016, cassedy_virus_2021, di_nardo_ten_2021}.

@@

@@

Currently, the detection and diagnosis of parasite infections rely on several laboratory methods in addition to clinical symptoms, clinical history, travel history, and geographic location of patient. serology-based assays, molecular-based assays and microscopy  \citep{ndao_diagnosis_2009}. The use of the microscopy for the morphology-based diagnosis of disease remains the reference standard in parasitology although identification and matching of the life-cycle stages, despite its low cost, is very challenging even with expert knowledge \citep{blasco-costa_molecular_2016}. LFAIs has been used successfully, overcoming many of the disadvantages of conventional microscopy by providing simple, quick, cost-effective, and sensitive and accurate identification of parasites, particularly in low-income countries, but are available for a relatively low number of organisms, low sensitivity and reports of cross-reactivity \citep{momcilovic_rapid_2019, robert-gangneux_epidemiology_2012, noauthor_false_nodate}.

@@

@@

\subsection{Sequencing}

\subsection{The role of bioinformatics}

\section{Metagenomics}

>>>>

In addition to the emergence of virulent pathogens, the rise of antimicrobial resistance (AMR) poses a major threat to human health around the world. In 2019 there were an estimated 4·95 million deaths associated with bacterial AMR \citep{murray_global_2022}. In 2017, the WHO released The Global Priority Pathogens (GPP) list \citep{world_health_organization_prioritization_2017} to guide discovery, research and development of new antibiotics for drug-resistant bacterial infections. Besides tuberculosis, the global priority due to being the most common and lethal airborne AMR disease worldwide today, responsible for 250 000 deaths each year, it includes 12 groups of pathogens in three priority categories:

@@

\subsection{Current standards for diagnostic in clinical microbiology}

The past few decades have seen a major revolution in the operation of microbial laboratories, driven by the development of molecular technologies and ways to make these accessible, namely amplification-based polymerase chain reaction (PCR), matrix-assisted laser desorption/ionisation - time of light (MALDI-TOF) and DNA-microarray-based hybridisation technology. These are used in conjunction with traditional techniques such as microscopy, culture and serology, not fully replacing them. Application of these methods differs by suspected infection type: bacterial, viral, fungal or parasitic.

@@

For patients with bacterial infections, the crucial steps are (1) to grow an isolate from a specimen, (2) identify its species, and (3) determine its pathogenic potential and test its susceptibility to antimicrobial drugs  \citep{didelot_transforming_2012}. Together this information facilitates the specific and rational treatment of patients. For public health purposes, knowledge also needs to be gained about (4) the relatedness of the pathogen to other strains of the same species to investigate transmission routes and enable the recognition of outbreaks \citep{foxman_choosing_2005}.

The current gold standard for bacterial pathogen identification in diagnostic microbiology laboratories involves the isolation of the pathogen through culture followed by biochemical testing, multi-step process that can take days to weeks before obtaining results, depending on the fastidiousness of the organism and if it can be cultured \citep{muhamad_rizal_advantages_2020, giuliano_guide_2019, muhamad_rizal_advantages_2020}. Although culture allows the identification of a wide variety of organisms, some pathogens can escape routine investigation due to strict metabolic necessities for growth or requirement for specific biochemical tests needed for their identification. Additionally, results will be obscured if a mixed culture is obtained, particularly if the cultures are obtained from sites with a microbiota, such as the gut and the skin, increasing the risk of contamination from normal flora, leading to false results \citep{giuliano_guide_2019}. After a successful growth in culture, Gram staining and MALDI-TOF mass spectometry is often used for identification with good accuracy as long as the pathogen is presented in the the coexisting database \citep{patel_maldi-tof_2015}. An alternate rapid identification method is PCR where nucleic acid fragments are detected through specific probes, being highly sensitive and specific, to the point where PCR may detect bacteria that are not viable after a patient has been treated for an infection \citep{scerbo_beyond_2016}. It's limited to the probes used, with multiplex PCR, an extension of PCR by using multiple primers, tried to address this issue by allowing for the identification of multiple bacteria and other important information such as the detection of antibiotic resistance genes or virulence genes \citep{giuliano_guide_2019}

Following identification, antibiotic-susceptibility testing is essential for guiding clinicians in selecting an appropriate treatment. Conventional detection methods of bacterial resistance, such as disc diffusion, antimicrobial gradient strip and broth microdilution, are widely used but results cannot be obtained earlier than 48 hours after receiving a sample, which may lead to prolonged use or overuse of broad-spectrum antibiotics \citep{benkova_antimicrobial_2020}. Similarly to bacterial identification, MALDI-TOF and PCR has been increasingly adopted as solutions with lower turnaround time, although no phenotipic information is retrieved, nor information on the minimum inhibitory concentration (MIC) for a given antibiotic.

@@

In the early 2000s, Multilocus sequence typing (MLST) was proposed as a portable, universal, and definitive method for characterising bacteria \citep{maiden_multilocus_2006}. Instead of enzyme restriction of bacteria DNA, separation of the restricted DNA bands using a pulsed-field electrophoresis chamber, followed by clonal assignment of bacteria based on PFGE banding patterns, MLST relies on the amplification through PCR sequences of internal fragments of house-keeping genes (usually 5 to 7), approximately 450-500 basepairs (bp) in size, followed by its the sequence, usually my Sanger methods. For each house-keeping gene, the different sequences present within a bacterial species are assigned as distinct alleles and, for each isolate, the alleles at each of the (usually) seven loci define the allelic profile or sequence type \citep{larsen_multilocus_2012}. As with PFGE, different schemes, defining what house-keeping gene fragments are used, are available depending of the species. Unlike PFGE, the provision of freely accessible, curated databases of MLST nucleotide sequence data enables the direct comparison of bacterial isolates, providing the basis of a common language for bacterial typing \citep{maiden_multilocus_2006}. So far, MLST schemes for 115 bacterial organisms having been published and made freely available (\url{https://pubmlst.org/organisms}, \cite{jolley_open-access_2018})

@@

Direct detection methods negate the need for virus propagation, detecting the virus directly from the suspect source through nucleic acid and immunological methods. PCR and reverse transcription-PCR are widely applied methods for the detection of both DNA and RNA viruses, respectively, driven by an increased awareness of the clinical value of, and demand for, prompt information about viral loads, viral sequence data, and potential antiviral resistance information \citep{cassedy_virus_2021}. Syndromic testing, utilising multiplex PCR to simultaneously amplify the nucleic acid of multiple targets in a single reaction, are now fully integrated into the standard testing practices of many clinical laboratories \citep{dien_bard_panels_2020}. Limitations of these assays include no detection of off-target pathogens, a lack of full susceptibility information, cost, and false-positive results.

Real-time quantitative PCR (qPCR) remains the front line tool in aetiological diagnosis, measuring the production of the target amplicon throughout the reaction and providing quantitative results with high specificity and sensibility, albeit with a significant cost due to sophisticated apparatus despite high-throughput systems being widely established \citep{cassedy_virus_2021}. Immunoassays employ singular-epitope specificity antibodies as the primary means to detect viruses within a sample, and provides as a much more cost-efficient alternative to nucleic acid detection \citep{cassedy_virus_2021}. One major application is seroprevalence assays, an essential technique for identifying patients who have been exposed to a virus (historical exposure), detect asymptomatic infection or evaluate vaccine efficacy  \citep{chan_determining_2021, bobrovitz_global_2021}. Lateral flow immunoassays (LFIA) are extensively used for detecting virus-associated protein directly from source through labelled antibodies binding to their cognate antigens, usually read by way of a colour change at a test line. Besides being very cost effective, LFAIs have a turnaround time of minutes and the colour change can be observed with the naked eye, therefore facilitating rapid diagnosis but its results are limited to semi-quantitative and it does not typically achieve sensitivity comparable to nucleic-acid detection \citep{estrela_lateral_2016, cassedy_virus_2021, di_nardo_ten_2021}.

@@

As with bacterial infections, typing through MLST is also performed for epidemiological studies, with public schemes being available for seven species: \textit{Aspergillus fumigatus}, \textit{Candida albicans}, \textit{Candida glabrata}, \textit{Candida krusei}, \textit{Candida tropicalis}, \textit{Geotrichum spp.} and \textit{Streptomyces spp} (\url{https://pubmlst.org/organisms}, \cite{jolley_open-access_2018}).

@@

Currently, the detection and diagnosis of parasite infections rely on several laboratory methods in addition to clinical symptoms, clinical history, travel history, and geographic location of patient. serology-based assays, molecular-based assays and microscopy  \citep{ndao_diagnosis_2009}. The use of the microscopy for the morphology-based diagnosis of disease remains the reference standard in parasitology although identification and matching of the life-cycle stages, despite its low cost, is very challenging even with expert knowledge \citep{blasco-costa_molecular_2016}. LFAIs has been used successfully, overcoming many of the disadvantages of conventional microscopy by providing simple, quick, cost-effective, and sensitive and accurate identification of parasites, particularly in low-income countries, but are available for a relatively low number of organisms, low sensitivity and reports of cross-reactivity \citep{momcilovic_rapid_2019, robert-gangneux_epidemiology_2012, rajput_false_2018}.

@@

Currently, public MLST schemes for for epidemiological studies of parasitic infections are available for six organisms: \textit{Blastocystis spp.},\textit{ Clonorchis sinensis}, \textit{Kudoa septempunctata}, \textit{Saprolegnia parasitica}, \textit{Treponema pallidum subsp. pallidum} and \textit{Tricohomas vaginalis} (\url{https://pubmlst.org/organisms}, \cite{jolley_open-access_2018}).

\subsection{Surveillance and infection prevention in public health}

@@

\subsection{Genomics}

\subsection{Metagenomics}

\section{The role of bioinformatics}
